### 2022-03-17 chapters/conclusions/conclusions.tex
new title

Our coupled hydro-epidemiological model reproduced the observed propagation of the SCTLD through the FRT by representing the transport of its causative agent within neutrally buoyant material driven by mean barotropic currents. It confirmed the result of previous ex-situ studies that showed evidence of waterborne transmission of the SCTLD with an average transmission time of the order of 10 days. Furthermore, by using coral resistance to the disease as a parameter, our model results suggested that, on average, corals had a rather low resistance against the disease. This confirmation of experimental results by model results illustrates the important potential of models in evaluating hypotheses derived from field observations on large scale systems. As such, models informed by and confronted against field knowledge \citep{foster2012connectivity} are a powerful tool to inform the management of complex ecosystems such as the FCR.

@@

\subsection*{Combining disease and larval connectivity}
All chapters dedicated to SCTLD rely on the concept of connectivity. In this context, connectivity negatively impacts coral reefs as it drives the propagation of disease agents. However, connectivity is actually a double-edged sword. Larval connectivity promotes the resilience of the reef network as it allows the recolonization of disturbed reefs following extreme events such as mass bleaching, hurricanes or disease outbreaks. Larval connectivity is therefore a valuable tool to inform coral reef protection and restoration \citep{botsford2009connectivity,mumby2011reserve}. Identifying reefs that are both good sources and sinks of larvae in the connectivity network indicates the optimal sites for coral outplants in the context of active restoration. Furthermore, knowledge about larval connectivity would allow us to build robust and sustainable networks of connected marine protected areas. As SLIM allows the computation of larval connectivity \citep{thomas2014numerical,frys2020fine,figueiredo2021global}, a possible application of the tools developed in this thesis is the combination of both types of connectivity.  This combined information would allow us to identify reefs that are good larval sources and have low vulnerability to the disease. A possible measure of larval supply to the network is the (weighted) out-degree of reefs in the larval connectivity network, as reefs with strong and numerous outgoing connections in the network are more likely to send many larvae to many connected reefs. Vulnerability to disease can be evaluated using the (weighted) in-degree of reefs in the disease connectivity network, as reefs with strong and numerous incoming connections are more likely to receive more disease agents from many connected reefs. Therefore, reefs with high restoration potential would correspond to reefs with large larval (weighted) out-degree and low disease (weighted) in-degree. This approach have been applied in \cite{holstein2022} and could be further developed in future studies using more complex connectivity metrics such as the PageRank index \citep{frys2020fine}.

>>>>

Our coupled hydro-epidemiological model reproduced the observed propagation of the SCTLD through the FRT by representing the transport of its causative agent within neutrally buoyant material driven by mean barotropic currents. It confirmed the results of previous ex-situ studies that showed evidence of waterborne transmission of the SCTLD with an average transmission time of the order of 10 days. Furthermore, by using coral resistance to the disease as a parameter, our model results suggested that, on average, corals had a rather low resistance against the disease. This confirmation of experimental results by model results illustrates the important potential of models to evaluate hypotheses derived from field observations on large scale systems. As such, models informed by and confronted against field knowledge \citep{foster2012connectivity} are a powerful tool to inform the management of complex ecosystems such as the FCR.

@@

\subsection*{Combining disease and larval connectivity}
All chapters dedicated to SCTLD rely on the concept of connectivity. In this context, connectivity negatively impacts coral reefs as it drives the propagation of disease agents. However, connectivity is actually a double-edged sword. Larval connectivity promotes the resilience of the reef network as it allows the recolonization of disturbed reefs following extreme events such as mass bleaching, hurricanes or disease outbreaks. Larval connectivity is therefore a valuable tool to inform coral reef protection and restoration \citep{botsford2009connectivity,mumby2011reserve}. Identifying reefs that are both good sources and sinks of larvae in the connectivity network indicates the optimal sites for coral outplants in the context of active restoration. Furthermore, knowledge about larval connectivity would allow us to build robust and sustainable networks of connected marine protected areas. As SLIM allows the computation of larval connectivity \citep{thomas2014numerical,frys2020fine,figueiredo2021global}, a possible application of the tools developed in this thesis is the combination of both types of connectivity.  This combined information would allow us to identify reefs that are good larval sources and have low vulnerability to the disease. A possible measure of larval supply to the network is the (weighted) out-degree of reefs in the larval connectivity network, as reefs with strong and numerous outgoing connections in the network are more likely to send many larvae to many connected reefs. Vulnerability to disease can be evaluated using the (weighted) in-degree of reefs in the disease connectivity network, as reefs with strong and numerous incoming connections are more likely to receive more disease agents from many connected reefs. Therefore, reefs with high restoration potential would correspond to reefs with large larval (weighted) out-degree and low disease (weighted) in-degree. This approach has been applied in \cite{holstein2022} and could be further developed in future studies using more complex connectivity metrics such as the PageRank index \citep{frys2020fine}.

### 2022-03-17 chapters/drto/article.tex
another fig modification
new title

Previous study of the spatio-temporal dynamics of SCTLD between West Palm Beach and Marathon showed that the epizootic propagated at a rate of $\sim$100 m/day to the north and $\sim$92 m/day to the south, hitting the Middle Florida Keys in 2017 \citep{muller2020spatial}. The disease was then reported in the Lower Keys in summer 2018 \citep{frrp2018}; near Key West in winter 2018 \citep{frrp2018}; on the easternmost boundary of the Marquesas subregion in October 2019 \citep{frrp2019}; and at its western end, approximately 11 km. west of halfmoon shoal, in August 2020 \citep{frrp2020}. SCTLD was observed propagating faster than 100 m/day within the Marquesas between October 2019 and August 2020 and would have reached the reefs of the Dry Tortugas (DRTO) by February 2021 if it had spread at a similarly high propagation rate \citep{kramer2019map}. These propagation rates were observed in a relatively dense shallow-water reef system and mostly occurred through transmission from neighbor to neighbor. Disease agents then had to cross 30 km ($\sim$20 miles) of open waters to reach the reefs of the DRTO. The disease was ultimately observed in May 2021 \citep{kramer2019map} in the southeast of the DRTO. To date, questions remain about the timing of the spread of the disease to the DRTO as well as the role played by hydrodynamics in the process.

@@

% === METHODS === %
\section{Methods}
Stony Coral Tissue Loss Disease data was obtained by using the Florida Reef Resilience Program's Disturbance Response Monitoring Report for 2019 and 2020 \citep{frrp2019,frrp2020} as well as the Atlantic and Gulf Rapid Reef Assessment (AGGRA) resources for SCTLD reported throughout the Caribbean, an open access spatial and temporal database curated by experts in Caribbean coral reef ecology \citep{kramer2019map}. Assumptions associated with SCTLD included the following: $(i)$ SCTLD is different from other tissue loss diseases previously reported on corals (i.e. white plague, white syndrome) based on its ecology and case definition \citep{noaa2018}, $(ii)$ SCTLD is transmitted, in part, by ocean currents \citep{aeby2019pathogenesis, muller2020spatial,eaton2021measuring} and $(iii)$ SCTLD was along the western most reef area of the Lower Florida Keys in October 2019, reached the western most part of the Marquesas in August 2020, and did not occur in DRTO until May 2021 based on curated data provided by the FRRP reports and the AGGRA open access database.

@@

We simulated the dispersal of disease agents throughout FCR between May 2018 and May 2021. As in \cite{dobbelaere2020coupled}, disease agents were modeled as passive particles transported by depth-averaged currents which constantly lost mass with a half-life of 30 days. We ran the particle transport model for every month between May 2018 and May 2021 and hence obtained monthly normalized connectivity matrices. An entry $C_{ij}$ of the connectivity matrix represents the probability that disease agents produced on sub-reef $i$ settle on sub-reef $j$. These sub-reefs were defined by extracting polygons from the “coral reefs and hard bottom” layer of the Unified Florida Reef Tract Map \citep{fwc2017unified} and further dividing them into 500 m $\times$ 500 m squares, generating a total of $16,823$ sub-reef polygons. Since the connectivity matrices had more than 200 million entries, they were not directly interpreted. Instead, they were represented as large networks where nodes corresponded to sub-reefs of FCR and edges corresponded to nonzero entries in the connectivity matrix. Graph-theory algorithms were then used to analyze the properties of the connectivity networks.

To quantify the connectivity between the Marquesas and the DRTO, we defined 3 sub-regions within the sub-reefs of the western part of FCR: $(i)$ the Marquesas, $(ii)$ the Western Marquesas and $(iii)$ the DRTO (Fig. \ref{fig:fig1_drto}d). The sub-region of the Marquesas was defined as all the sub-reefs whose centroid is located between $-82.21^\circ$ and $-82.05^\circ$ longitude; the Western Marquesas included all sub-reefs between $-82.73^\circ$ and $-82.21^\circ$; and the DRTO were composed of  all the sub-reefs in our domain located west of $-82.73^\circ$ longitude. We evaluated the exchanges between two given sub-regions by identifying all edges connecting any sub-reef from one sub-region to any sub-reef from the other sub-region in our monthly connectivity networks. We then defined a connectivity indicator by multiplying the number of these direct edges by their mean connection probability. This indicator was computed monthly between May 2018 and May 2021 to evaluate the connectivity $(i)$ from the Marquesas to the DRTO, $(ii)$ from the Western Marquesas to the DRTO and $(iii)$ from the Marquesas to the Western Marquesas.

@@

\begin{figure}
    \centering
    \includegraphics[width=.9\textwidth]{chapters/drto/figures/fig4.png}
    \caption{ (a) Vulnerability indicator value of all sub-reefs of the DRTO in November 2020. Eastern reefs of the DRTO appear most vulnerable to the disease. Locations where the SCTLD was observed in the DRTO are indicated by red stars (b) Source indicator value in November 2020 for every sub-reef of the Marquesas and the Western Marquesas. Northwestern reefs of the Western Marquesas and southern reefs between the Marquesas and the Western Marquesas show a higher transmission potential to the DRTO. Main islands are shown in dark grey, separation between the Marquesas and the Western Marquesas is indicated by a dashed line.}
    \label{fig:fig4_drto}
\end{figure}

>>>>

Previous study of the spatio-temporal dynamics of SCTLD between West Palm Beach and Marathon showed that the epizootic propagated at a rate of $\sim$100 m/day to the north and $\sim$92 m/day to the south, hitting the Middle Florida Keys in 2017 \citep{muller2020spatial}. The disease was then reported in the Lower Keys in summer 2018 \citep{frrp2018}; near Key West in winter 2018 \citep{frrp2018}; on the easternmost boundary of the Marquesas subregion in October 2019 \citep{frrp2019}; and at its western end, approximately 11 km west of halfmoon shoal, in August 2020 \citep{frrp2020}. SCTLD was observed propagating faster than 100 m/day within the Marquesas between October 2019 and August 2020 and would have reached the reefs of the Dry Tortugas (DRTO) by February 2021 if it had spread at a similarly high propagation rate \citep{kramer2019map}. These propagation rates were observed in a relatively dense shallow-water reef system and mostly occurred through transmission from neighbor to neighbor. Disease agents then had to cross 30 km ($\sim$20 miles) of open waters to reach the reefs of the DRTO. The disease was ultimately observed in May 2021 \citep{kramer2019map} in the southeast of the DRTO. To date, questions remain about the timing of the spread of the disease to the DRTO as well as the role played by hydrodynamics in the process.

@@

% === METHODS === %
\section{Methods}
Stony Coral Tissue Loss Disease data was obtained by using the Florida Reef Resilience Program's Disturbance Response Monitoring Report for 2019 and 2020 \citep{frrp2019,frrp2020} as well as the Atlantic and Gulf Rapid Reef Assessment (AGGRA) resources for SCTLD reported throughout the Caribbean, an open access spatial and temporal database curated by experts in Caribbean coral reef ecology \citep{kramer2019map}. Assumptions associated with SCTLD included the following: (\textit{i}) SCTLD is different from other tissue loss diseases previously reported on corals (i.e. white plague, white syndrome) based on its ecology and case definition \citep{noaa2018}, (\textit{ii}) SCTLD is transmitted, in part, by ocean currents \citep{aeby2019pathogenesis, muller2020spatial,eaton2021measuring} and (\textit{iii}) SCTLD was along the western most reef area of the Lower Florida Keys in October 2019, reached the western most part of the Marquesas in August 2020, and did not occur in DRTO until May 2021 based on curated data provided by the FRRP reports and the AGGRA open access database.

@@

We simulated the dispersal of disease agents throughout FCR between May 2018 and May 2021. As in \cite{dobbelaere2020coupled}, disease agents were modeled as passive particles transported by depth-averaged currents which constantly lost mass with a half-life of 30 days. We ran the particle transport model for every month between May 2018 and May 2021 and hence obtained monthly normalized connectivity matrices. An entry $C_{ij}$ of the connectivity matrix represents the probability that disease agents produced on sub-reef $i$ settle on sub-reef $j$. These sub-reefs were defined by extracting polygons from the “coral reefs and hard bottom” layer of the Unified Florida Reef Tract Map \citep{fwc2017unified} and further dividing them into 500 m $\times$ 500 m squares, generating a total of 16,823 sub-reef polygons. Since the connectivity matrices had more than 200 million entries, they were not directly interpreted. Instead, they were represented as large networks where nodes corresponded to sub-reefs of FCR and edges corresponded to nonzero entries in the connectivity matrix. Graph-theory algorithms were then used to analyze the properties of the connectivity networks.

To quantify the connectivity between the Marquesas and the DRTO, we defined 3 sub-regions within the sub-reefs of the western part of FCR: (\textit{i}) the Marquesas, (\textit{ii}) the Western Marquesas and (\textit{iii}) the DRTO (Fig. \ref{fig:fig1_drto}d). The sub-region of the Marquesas was defined as all the sub-reefs whose centroid is located between $-82.21^\circ$ and $-82.05^\circ$ longitude; the Western Marquesas included all sub-reefs between $-82.73^\circ$ and $-82.21^\circ$; and the DRTO were composed of  all the sub-reefs in our domain located west of $-82.73^\circ$ longitude. We evaluated the exchanges between two given sub-regions by identifying all edges connecting any sub-reef from one sub-region to any sub-reef from the other sub-region in our monthly connectivity networks. We then defined a connectivity indicator by multiplying the number of these direct edges by their mean connection probability. This indicator was computed monthly between May 2018 and May 2021 to evaluate the connectivity (\textit{i}) from the Marquesas to the DRTO, (\textit{ii}) from the Western Marquesas to the DRTO and (\textit{iii}) from the Marquesas to the Western Marquesas.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/drto/figures/fig4.png}
    \caption{ (a) Vulnerability indicator value of all sub-reefs of the DRTO in November 2020. Eastern reefs of the DRTO appear most vulnerable to the disease. Locations where the SCTLD was observed in the DRTO are indicated by red stars (b) Source indicator value in November 2020 for every sub-reef of the Marquesas and the Western Marquesas. Northwestern reefs of the Western Marquesas and southern reefs between the Marquesas and the Western Marquesas show a higher transmission potential to the DRTO. Main islands are shown in dark grey, separation between the Marquesas and the Western Marquesas is indicated by a dashed line.}
    \label{fig:fig4_drto}
\end{figure}

### 2022-03-17 chapters/drto/drto.tex
updated ref for onset
new title

\item \textbf{Dobbelaere, T.}, Holstein, D. M., Muller, E. M., Gramer, L. J., McEachron L., Williams, S. D., $\&$ Hanert, E. Connecting the dots: Transmission of stony coral tissue loss disease from the Marquesas to the Dry Tortugas. accepted for publication in \textit{Frontiers in Marine Science}.\\ \href{https://doi.org/10.3389/fmars.2022.778938}{doi: 10.3389/fmars.2022.778938}.

@@

\begin{abstract}
    For the last seven years, Florida's Coral Reef (FCR) has suffered from widespread and severe coral loss caused by stony coral tissue loss disease (SCTLD). First observed off the coast of Miami-Dade county in 2014, the outbreak has since spread throughout the entirety of FCR and some areas of  the Caribbean. However, the propagation of the disease through FCR seemed to slow down when it reached the western end of the Marquesas in August 2020. Despite being present about 30 km ($\sim$20 miles) from the Dry Tortugas (DRTO), SCTLD was not reported in this area before May 2021. As SCTLD transmission is likely to be waterborne, here we suggest that this apparently delayed propagation is related to eddy activity near the DRTO under the influence of the Loop Current/Florida Current system. To quantify the impact of the local ocean circulation on the spread of SCTLD from the Marquesas and the DRTO, we evaluated the hydrodynamic-predicted connectivity between these two regions using a high-resolution hydro-epidemiological model between May 2018 and May 2021. Our results suggest that the Marquesas and the DRTO were not connected during February-October 2020 and January-May 2021. These periods coincided with either the occurrence of Tortugas gyres and mean circulation with an eastward component between the Marquesas and the DRTO or the presence of southward currents. Our results suggest that disease agents probably reached the DRTO in November 2020 and that they most likely originated from southern or northwestern reefs of the Marquesas. This study provides novel insight into the role played by the hydrodynamics in the spread of SCTLD within the western-most edge of FCR, and in propagating the disease to uninfected locations.
\end{abstract}

>>>>

\item \textbf{Dobbelaere, T.}, Holstein, D. M., Muller, E. M., Gramer, L. J., McEachron L., Williams, S. D., \& Hanert, E. (2022). Connecting the dots: Transmission of stony coral tissue loss disease from the Marquesas to the Dry Tortugas. \textit{Frontiers in Marine Science}, 9. \href{https://www.frontiersin.org/article/10.3389/fmars.2022.778938}{doi: 10.3389/fmars.2022.778938}.

@@

\begin{abstract}
    For the last seven years, Florida's Coral Reef (FCR) has suffered from widespread and severe coral loss caused by the stony coral tissue loss disease (SCTLD). First observed off the coast of Miami-Dade county in 2014, the outbreak has since spread throughout the entirety of FCR and some areas of  the Caribbean. However, the propagation of the disease through FCR seemed to slow down when it reached the western end of the Marquesas in August 2020. Despite being present about 30 km ($\sim$20 miles) from the Dry Tortugas (DRTO), SCTLD was not reported in this area before May 2021. As SCTLD transmission is likely to be waterborne, here we suggest that this apparently delayed propagation is related to eddy activity near the DRTO under the influence of the Loop Current/Florida Current system. To quantify the impact of the local ocean circulation on the spread of SCTLD from the Marquesas and the DRTO, we evaluated the hydrodynamic-predicted connectivity between these two regions using a high-resolution hydro-epidemiological model between May 2018 and May 2021. Our results suggest that the Marquesas and the DRTO were not connected during February-October 2020 and January-May 2021. These periods coincided with either the occurrence of Tortugas gyres and mean circulation with an eastward component between the Marquesas and the DRTO or the presence of southward currents. Our results suggest that disease agents probably reached the DRTO in November 2020 and that they most likely originated from southern or northwestern reefs of the Marquesas. This study provides novel insight into the role played by the hydrodynamics in the spread of SCTLD within the western-most edge of FCR, and in propagating the disease to uninfected locations.
\end{abstract}

### 2022-03-17 chapters/intro/intro.tex
new title

The calcifying process of reef building corals heavily depends on their symbiosis with the microalgae zooxanthellae. These unicellular symbionts convert sunlight and carbon dioxide into organic carbon and oxygen, providing the corals with most of the energy needed to meet their metabolic demands \citep{muscatine1977reef}. However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the mechanisms maintaining this association, resulting in the expulsion of the endosymbionts \citep{hoegh2007coral,van2022global}. This process, called bleaching, causes the coral to loose their pigmentation and reveals their underlying white skeleton \citep{baker2008climate}. Bleached corals are then physiologically and nutritionally compromised, and prolonged bleaching over several months leads to high coral mortality \citep{hughes2018spatial}. Decline in coral health and cover pushes the reef ecosystem to a macroalgal-dominated state \citep{hughes2003climate,mumby2007thresholds}. Once a "tipping point" is exceeded, returning to a coral-dominated state becomes difficult \citep{mumby2007thresholds,graham2015predicting}.

@@

Due to the combined damages from of all these stressors, it is almost impossible to return reefs to their past configurations \citep{hughes2017coral}. Instead, the challenge is now to maintain the biological functions of coral reefs through quickly changing environmental conditions. This will require to rapidly address greenhouse gas emissions, as well as local actions to boost the capacity of coral reef ecosystems to survive climate change \citep{hughes2003climate,knowlton2008shifting,graham2015predicting,kleypas2021designing}. The aim of the present dissertation was therefore to use modeling tools to better understand the drivers of coral demise and hence inform such local actions to support the management of Florida's Coral Reef, the third largest barrier reef of the world.

@@

The latest major outbreak affecting coral reefs of the Caribbean is the SCTLD \citep{noaa2018}. First observed off the coasts of Miami in 2014 during the PMDDP \citep{precht2016unprecedented}, the disease has since spread throughout the entire Florida Reef Tract and numerous territories of the Caribbean \citep{alvarez2019rapid, kramer2019map, estrada2021effects} (Fig. \ref{intro:propagation}). The outbreak affects at least 24 species of scleractinian coral and often results in whole colony mortality \citep{precht2016unprecedented, walton2018impacts}. Briefly, the gross morphology of SCTLD can be focal or multifocal, with locally extensive to diffuse areas of acute to subacute tissue loss distributed basally, peripherally, or both. In some cases, tissues bordering areas of chronic tissue loss show indistinct bands of pallor, progressing to normal pigmentation away from the denuded skeleton (Fig. \ref{intro:sctld}). The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution suggest that SCTLD might be the largest coral disease outbreak on record in Florida. Although the causative agent of the disease remains unknown, there is evidence of waterborne disease transmission \citep{aeby2019pathogenesis,eaton2021measuring,meiling2021variable} and recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/hurricane_strikes.jpg}
    \caption{Total number of major hurricanes strikes in Florida between 1900-2010. Source: National Oceanographic and Atmospheric Administration, National Hurricane Center \url{https://www.nhc.noaa.gov/climo/}}
    \label{inro:landfall}
\end{figure}

@@

The aforementioned large-scale ocean circulation significantly impacts transport processes in the FRT. The modeling tools developed in this thesis should therefore be able to capture them accurately. Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation features. For instance, small-scale flow features such as recirculation eddies around reefs and islands have been shown to strongly influence larval dispersal by increasing their local retention over the reefs \citep{figueiredo2013synthesizing}. Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}. A reference ocean model of the FRT is the 3D HYbrid Coordinate Ocean Model (HYCOM, \citealp{chassignet2007hycom}), or more specifically the nested high resolution model FKEYS-HYCOM, that covers the Florida Straits, South Florida and Keys with a $\sim$900 m resolution \citep{kourafalou2012florida}. This model is embedded within the coarser GOM-HYCOM, that models the whole GoM and Florida Straits using data assimilation with a resolution of $\sim$3.5 km \citep{cummings2005operational,prasad2007upper}. However as a structured-grid model, the finer resolution of HYCOM is limited by computational resources and the nested grid approach lacks the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}. In contrast, unstructured-mesh models offer a well-suited solution to capture the multi-scale nature of flow features in the FRT in a computationally tractable way. They allow to locally refine the model resolution to $\sim~100$ m close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.

>>>>

The calcifying process of reef building corals heavily depends on their symbiosis with the microalgae zooxanthellae. These unicellular symbionts convert sunlight and carbon dioxide into organic carbon and oxygen, providing the corals with most of the energy needed to meet their metabolic demands \citep{muscatine1977reef}. However, adverse environmental conditions such as elevated sea surface temperature can cause damage to the mechanisms maintaining this association, resulting in the expulsion of the endosymbionts \citep{hoegh2007coral,van2022global}. This process, called bleaching, causes the corals to loose their pigmentation and reveals their underlying white skeleton \citep{baker2008climate}. Bleached corals are then physiologically and nutritionally compromised, and prolonged bleaching over several months leads to high coral mortality \citep{hughes2018spatial}. Decline in coral health and cover pushes the reef ecosystem to a macroalgal-dominated state \citep{hughes2003climate,mumby2007thresholds}. Once a "tipping point" is exceeded, returning to a coral-dominated state becomes difficult \citep{mumby2007thresholds,graham2015predicting}.

@@

Due to the combined damages from of all these stressors, it is almost impossible to return reefs to their past configurations \citep{hughes2017coral}. Instead, the challenge is now to maintain the biological functions of coral reefs through quickly changing environmental conditions. This will require to rapidly address greenhouse gas emissions, as well as local actions to boost the capacity of coral reef ecosystems to survive climate change \citep{hughes2003climate,knowlton2008shifting,graham2015predicting,kleypas2021designing}. The aim of the present dissertation is therefore to use modeling tools to better understand the drivers of coral demise and hence inform such local actions to support the management of Florida's Coral Reef, the third largest barrier reef of the world.

@@

The latest major outbreak affecting coral reefs of the Caribbean is the SCTLD \citep{noaa2018}. First observed off the coasts of Miami in 2014 during the PMDDP \citep{precht2016unprecedented}, the disease has since spread throughout the entire FRT and numerous territories of the Caribbean \citep{alvarez2019rapid, kramer2019map, estrada2021effects} (Fig. \ref{intro:propagation}). The outbreak affects at least 24 species of scleractinian coral and often results in whole colony mortality \citep{precht2016unprecedented, walton2018impacts}. Briefly, the gross morphology of SCTLD can be focal or multifocal, with locally extensive to diffuse areas of acute to subacute tissue loss distributed basally, peripherally, or both. In some cases, tissues bordering areas of chronic tissue loss show indistinct bands of pallor, progressing to normal pigmentation away from the denuded skeleton (Fig. \ref{intro:sctld}). The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution suggest that SCTLD might be the largest coral disease outbreak on record in Florida. Although the causative agent of the disease remains unknown, there is evidence of waterborne disease transmission \citep{aeby2019pathogenesis,eaton2021measuring,meiling2021variable} and recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/hurricane_strikes.jpg}
    \caption{Total number of major hurricanes strikes in Florida between 1900 and 2010. Source: National Oceanographic and Atmospheric Administration, National Hurricane Center \url{https://www.nhc.noaa.gov/climo/}}
    \label{inro:landfall}
\end{figure}

@@

The aforementioned large-scale ocean circulation significantly impacts transport processes in the FRT. The modeling tools developed in this thesis should therefore be able to capture them accurately. Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation features. For instance, small-scale flow features such as recirculation eddies around reefs and islands have been shown to strongly influence larval dispersal by increasing their local retention over the reefs \citep{figueiredo2013synthesizing}. Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}. A reference ocean model of the FRT is the 3D HYbrid Coordinate Ocean Model (HYCOM, \citealp{chassignet2007hycom}), or more specifically the nested high resolution model FKEYS-HYCOM, that covers the Florida Straits, South Florida and Keys with a $\sim$900 m resolution \citep{kourafalou2012florida}. This model is embedded within the coarser GOM-HYCOM, that models the whole GoM and the Straits of Florida using data assimilation with a resolution of $\sim$3.5 km \citep{cummings2005operational,prasad2007upper}. However as a structured-grid model, the finer resolution of HYCOM is limited by computational resources and the nested grid approach lacks the flexibility required to adapt to the complex topology of coastal regions \citep{fringer2019future}. In contrast, unstructured-mesh models offer a well-suited solution to capture the multi-scale nature of flow features in the FRT in a computationally tractable way. They allow to locally refine the model resolution to $\sim$100 m close to reefs and island while covering a domain large enough to capture the impact of the FC \citep{lambrechts2008multi,frys2020fine}.

### 2022-03-17 chapters/irma/irma.tex
updated ref for onset
new title

This chapter is based on the following article:
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%
\item \textbf{Dobbelaere, T.}, Curcic, M., Le Hénaff, M., & Hanert, E. (2022). Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes. \textit{Ocean Modelling}, 101947. \href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.1016/j.ocemod.2022.101947}.
\end{list}

\begin{abstract}
    The intensity of major tropical cyclones has increased during the past decade. Their effect is particularly acute in coastal areas where they cause extensive damage leading to an influx of debris, sediments and waste to the sea. However, most operational coastal ocean models do not represent heavy-wind transport processes correctly \modif{if} the hydrodynamics is not coupled with the wind-generated waves. This may lead to significant errors in ocean simulations under tropical cyclone conditions. Here, we investigate current-wave interactions during a major hurricane and assess their impact on transport processes. We do that by coupling the unstructured-mesh coastal ocean model SLIM with the spectral wave model SWAN, and applying it to the Florida Reef Tract during Hurricane Irma (September 2017). We show that the coupled model successfully reproduces the wave behavior, the storm surge and the ocean currents during the passage of the hurricane. We then use the coupled and uncoupled wave-current model to simulate the transport of passive drifters. We show that the wave radiation stress gradient alone can lead to changes of up to 1 m/s in the modeled currents, which in turn leads to differences of up to 5 km in the position of drifting material over the duration of the hurricane. The Stokes drift however appears to cause deflections up to 4 times larger and hence dominates wave-induced transport. Wave-current interactions therefore strongly impact the transport of drifting material such as sediments and debris in the aftermath of a hurricane. They should thus be taken into account in order to correctly assess its overall impact.
\end{abstract}

>>>>

This chapter is based on the following article:
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{0.23in}%
\setlength{\listparindent}{-0.23in}%
\setlength{\itemindent}{-0.23in}%
\setlength{\parsep}{\parskip}%
}%
\item \textbf{Dobbelaere, T.}, Curcic, M., Le Hénaff, M., \& Hanert, E. (2022). Impacts of Hurricane Irma (2017) on wave-induced ocean transport processes. \textit{Ocean Modelling}, 101947. \href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.1016/j.ocemod.2022.101947}.
\end{list}

\begin{abstract}
    The intensity of major tropical cyclones has increased during the past decade. Their effect is particularly acute in coastal areas where they cause extensive damage leading to an influx of debris, sediments and waste to the sea. However, most operational coastal ocean models do not represent heavy-wind transport processes correctly if the hydrodynamics is not coupled with the wind-generated waves. This may lead to significant errors in ocean simulations under tropical cyclone conditions. Here, we investigate current-wave interactions during a major hurricane and assess their impact on transport processes. We do that by coupling the unstructured-mesh coastal ocean model SLIM with the spectral wave model SWAN, and applying it to the Florida Reef Tract during Hurricane Irma (September 2017). We show that the coupled model successfully reproduces the wave behavior, the storm surge and the ocean currents during the passage of the hurricane. We then use the coupled and uncoupled wave-current model to simulate the transport of passive drifters. We show that the wave radiation stress gradient alone can lead to changes of up to 1 m/s in the modeled currents, which in turn leads to differences of up to 5 km in the position of drifting material over the duration of the hurricane. The Stokes drift however appears to cause deflections up to 4 times larger and hence dominates wave-induced transport. Wave-current interactions therefore strongly impact the transport of drifting material such as sediments and debris in the aftermath of a hurricane. They should thus be taken into account in order to correctly assess its overall impact.
\end{abstract}
